\thetitle

### 2020-11-20 writing/rational.tex
checkpoint

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug.

Equipped with these preliminary definitions, we can precisely define what it means
for blame to useful in the context of a gradual type system for a specific
type-level mistake independently of a program's type annotations. 
\begin{quote}
\it
Given a program $\system$, the natural blame mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing natural blame trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful natural blame trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural blame
trails.\\
\end{quote}

While these definitions describe how a rational programmer might gauge whether it
pays off to heed blame assignments during debugging, they do not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
\end{quote}
Similar to natural blame we determine how successful the exception
migration is based on the success of natural exception trails: 
\begin{quote}
\it
Given a program $\system$, the natural exception mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing
  natural exception trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful
  natural exception trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
exception trails.\\
\end{quote}

Putting the pieces together, we can now define when blame 
is more useful than mere exceptions in the context of a natural semantics.
\begin{quote}
\it Natural blame  is more useful
  than natural exceptions for $\system$ if either natural blame is
  always useful for $\system$ when natural exception is sometimes or
  never useful, or natural blame is sometimes useful for $\system$
  when natural exception is never useful.
\end{quote}

@@

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when the transient-last and transient-first blame modes are more
useful than exceptions in the context of a natural semantics.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of natural?

\item[$Q_2$] Is first blame useful in the context of transient?

\item[$Q_3$] Is last blame useful in the context of transient?

@@

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X more useful than blame for Y where X, Y is natural,
  erasure or natural?
\end{itemize}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by analyzing the natural blame and
natural exception trails in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select one of the interesting debugging
scenarios from section~\ref{sec:mutate}.
Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
same roots once more following the natural-exception programmer. 
Next, based on the collected data, we calculate the
success/failure results for the trails to determine whether blame
is more useful than exceptions for this mutant. Finally, we
obtain a positive answer for
$Q_1$ when the percentage of mutants where blame is more useful
than exceptions is greater than the percentage of mutants where exceptions
are more useful
than blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
modes of the rational programmer. For $Q_*$, we also proceed in a similar
manner except that we use modes 
of the rational programmer from different semantics. For instance to
decide whether blame for natural more useful than blame for
transient, we compare the usefulness of
the natural blame mode with that of the transient first and transient last
modes for all our mutants. 
Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.

>>>>

When the buggy untyped component of a program is replaced by the typed counterpart,
the type checker fails if this component causes the impedance mismatch. Hence, a
trail that ends at an ill-typed scenarios successfully pinpoints the location of
bugs. 
\begin{quote}
\it A natural blame trail $\conf_0,...\conf_n$ in a lattice $\lattice{\system}$ is
\emph{successful} iff its last scenario $\conf_n$ does not type check.  A natural
blame trail $\conf_0,..,\conf_n$ in a lattice $\lattice{\system}$ is \emph{failing}
iff $\conf_n$ type checks and the trail cannot be extended further.
\end{quote}
That is failing natural blame trails are those that have not reached a scenario that
reveals the bug at compile time and their last scenario does not produce blame. Thus
the rational programmer receives no hints from the gradual type system on how to
continue the search for the bug. In contrast, if the rational programmer
knows that for a program  $\system$
the lattice $\lattice{\system}$ contains only successful natural blame
trails, then the rational programmer can rely on natural blame while debugging
a type-level mistakes in $\system$ independently of $\system$'s type
annotations.

While the absence of failing natural blame trails in $\lattice{\system}$
imply that
pays off to heed blame assignments during debugging $\system$, it does not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
\end{quote}

Putting the two modes together, we can now compare the usefulness of blame 
and mere exceptions.
\begin{quote}
\it 
  Given a program $\system$, natural exceptions are  \emph{at least as useful} as 
  natural blame for $\system$ iff  for every interesting scenario 
  $\conf_0$ in $\lattice{\system}$ when the natural blame trail 
  that starts at $\conf_0$ is successful then the natural exception trail that
  starts at $\conf_0$ is also successful.
\end{quote}
\indent Of course if natural exceptions are not at least as useful as
natural blame for $\system$ then we have evidence that blame 
is \emph{valuable} for the rational programmer in the context of natural.

@@

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when transient exceptions are at least as useful as the 
transient-last and transient-first blame.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable in the context of natural?

\item[$Q_2$] Is first blame valuable in the context of transient?

\item[$Q_3$] Is last blame valuable in the context of transient?

@@

Furthermore, our experiment can compare the relative usefulness of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X at least as useful as blame for Y where X, Y is natural,
  erasure or natural?
\end{itemize}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by analyzing the natural blame and
natural exception trails in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select one of the interesting debugging
scenarios from section~\ref{sec:mutate}.
Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
same roots once more following the natural-exception programmer. 
Next, based on the collected data, we calculate the
success/failure results for the trails to determine whether exceptions are
as useful as blame for this mutant. Finally, we
obtain a positive answer for
$Q_1$ when there are mutants where natural exceptions are not as useful 
as natural blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
modes of the rational programmer. For $Q_*$, we also proceed in a similar
manner except that we use modes 
of the rational programmer from different semantics. For instance to
decide whether blame for natural is as useful as blame for
transient, we compare the success of
natural blame trails with that of the transient first and transient last
blame trails and we determine if one mode is as useful as another. 
Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.

### 2020-11-20 writing/results.tex
checkpoint

\begin{figure*}
  \centering
  \includegraphics[width=0.7\textwidth]{./plots/usefulness-table}
  \caption{Usefulness: Each cell plots the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).}
  \label{fig:usefulness-table}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=0.95\textwidth]{./plots/bt-lengths-table}
  \caption{Programmer effort: Each cell depicts the distribution of trail lengths for a given mode and benchmark. The leftmost bar of every distribution shows the proportion of trails with length zero, and each bar to the right shows the same for length one, two, three, and so on.}
  \label{fig:effort-table}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
  \caption{Adds-value comparisons: Each row depicts the percentage of mutants for which that row's mode adds value over each other mode (in the columns). The diagonals are ommitted because they compare a mode with itself.}
  \label{fig:avo-matrix}
\end{figure}

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=0.7\textwidth]{./plots/usefulness-table}
  \caption{Usefulness: Each cell depicts the proportion of a benchmark's mutants for which a mode is always useful (green), sometimes useful (yellow), and never useful (red).}
  \label{fig:usefulness-table}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=0.95\textwidth]{./plots/bt-lengths-table}
  \caption{Programmer effort: Each cell depicts the distribution of trail
  lengths for a given mode and a benchmark's mutants starting from trails
  of length 0.}
  \label{fig:effort-table}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{./plots/avo-matrix}
  \caption{Usefulness comparisons: Each cell depicts the percentage of a
  benchmark's mutants for which the row mode is more useful than the column mode.}
  \label{fig:avo-matrix}
\end{figure}

### 2020-11-20 writing/sample.tex
checkpoint

As outlined in section~\ref{sec:rational}, our experimental questions
require that we answer first whether a mode of the rational programmer is
useful or not for all the interesting scenarios of a mutant $\system$.
Given the binary nature of this basic question, we can rely on random
sampling of interesting scenarios from $\lattice{\system}$ to obtain an
approximate answer.  Thus $\lattice{\system}$ becomes the first stratum
for our sampling; with a sufficiently large sample, we can obtain from the
sample an estimate of the usefulness of a mode for the entire lattice.
The estimate tells us with some confidence whether the mode is useful for all trails in the lattice (if all of the samples yield positive results) or none of them.
In fact, the estimate can say neither if some of our samples yield positive results and others yield negative ones; in that case we can say no more than that the mode is sometimes useful.
We select a sample size of 96 to obtain 95\% confidence and a 5\% margin of error for our estimates
(using standard statistical methods, the choice does not depend on the population size for sufficiently large populations).
For lattices that have less than 96 interesting scenarios, we select them all.

>>>>

As outlined in section~\ref{sec:rational}, our experimental questions
boil down whether a mode of the rational programmer is
as useful as another mode for all the interesting scenarios of a mutant $\system$.
Given the binary nature of this basic question, we can rely on random
sampling of interesting scenarios from $\lattice{\system}$ to obtain an
approximate answer.  Thus $\lattice{\system}$ becomes the first stratum
for our sampling; with a sufficiently large sample, we can obtain from the
sample an estimate of the usefulness of a mode for the entire lattice.
The estimate tells us with some confidence whether the mode is useful for all trails in the lattice (if all of the samples yield positive results) or none of them.
In fact, the estimate can say neither if some of our samples yield positive results and others yield negative ones; in that case we can say no more than that the mode is sometimes useful.
We select a sample size of 96 to obtain 95\% confidence and a 5\% margin of error for our estimates
(using standard statistical methods, the choice does not depend on the population size for sufficiently large populations).
For lattices that have less than 96 interesting scenarios, we select them all.
